---
title: "Varying formative periods using Gumbel distribution"
author: "Nicolas Restrepo"
date: "4/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

```{r}
library(tidyverse)
library(extraDistr)
library(broom)
theme_set(theme_minimal())
```

I am going to explore the model we have but I will define the shape and length of formative periods using the Gumbel distribution. This distribution has a long right tail and, therefore, it can be used to describe events that are unlikely. I think it works here because it allows older agents to change but they have a small probability to do so. 

## Gumbel Distribution

I will run build density plots for the Gumbel distribution and then I will standardize the values so that they lie between 0 and 1, expressing probabilities. This is what the resulting distributions look like: 

```{r}
# Low standard deviation
y_lsd <- dgumbel(seq(0, 100, by = 1), mu = 20, sigma = 2)
y_lsd_scaled <- (y_lsd - min(y_lsd))/(max(y_lsd)-min(y_lsd))

# mid standard deviation
y_msd <- dgumbel(seq(0, 100, by = 1), mu = 20, sigma = 5)
y_msd_scaled <- (y_msd - min(y_msd))/(max(y_msd)-min(y_msd))    

# high standard deviation
y_hsd <- dgumbel(seq(0, 100, by = 1), mu = 20, sigma = 10)
y_hsd_scaled <- (y_hsd - min(y_hsd))/(max(y_hsd)-min(y_hsd))

df_gumbel <- tibble(age = rep(0:100, 3), 
                    p_learn = c(y_lsd_scaled, 
                                y_msd_scaled, 
                                y_hsd_scaled), 
                    type = rep(c("low_variance", 
                             "mid_variance", 
                             "high_variance"), each = 101))

df_gumbel %>% 
  ggplot(aes(x = age, 
             y = p_learn, 
             color = type)) +
  geom_line() + 
  labs(title = "Gumbel Distributions", 
       y = "Pr. Learning", 
       x = "Age",
       color = "Type")

```

In my opinion, this looks more realistic than what we had before. 

## Exploring the model

I think that exploring whether restricted formative windows aid with survival in certain environments is interesting but it's an avenue that has been widely explored by biologists. In fact, the whole argument is that formative windows develop in response to noisy and unreliable environments. We wouldn't be arguing anything new should we find evidence for that. 

I think the more interesting question for sociologists is whether the shape of formative windows changes the rhythm and nature of cultural change. To explore this, I am going to ask three questions from this ABM: 

1) *When the environment changes, how quickly do agents migrate to the favored trait?* This can give us an idea of the speed of cultural change. 

2) *How much consensus is there among agents across the run?* This helps us get at how much the population allows for the coexistence of diverging views. 

3) *How much of that consensus is made up of the favored trait?* This gets at how responsive the population is to environmental changes. 

I am going to run the model using the three different specifications for the varying windows shown above and I will compare the results. 

```{r}
# Function for the model
varying_windows_model <- function(N,
                                  t_max,
                                  b,
                                  bonus_fitness,
                                  bonus_survival, 
                                  num_rows, 
                                  p_soc_learn, 
                                  sigma_open, 
                                  sims, 
                                  mu_open) {
  # Empty list 
  all_results <- list()
  
  for (j in 1:sims) {
    #### Set-up ####
    # Repository (enough rows; plus one column for age)
    population <- matrix(NA, nrow= num_rows, 
                         ncol = t_max+2)
    #First traits randomly allocated
    population[1:N, 2] <- sample(c(1:4), 
                                 size = N, 
                                 replace = T)
    # Draw age as integer
    age <- runif(n = N, min = 1, max = 100) %>% as.integer()
    # Add age to our matrix 
    population[1:N,1] <- age
    # Initial state of the world
    state_world <- sample(c(1:4), 1)
    # An empty vector that records whether the world changes
    world_changes <- rep(NA, t_max)
    # Repository  for keeping track of our results
    df_results <- matrix(NA, nrow = t_max, ncol = 8)
    
    
    #### Loop ####
    for(i in 1:t_max) {
      # Notify me if everyone dies!
      if(length(which(!is.na(population[,1])))==0) {
        print("Everyone died!")
        break
      }
      # Keep track of last agent
      last_agent <- which(!is.na(population[,1]))[length(which(!is.na(population[,1])))]
      
      ### 1. Learning ####
      # Probability of learning
      x <- population[which(!is.na(population[,1])),1] %>% 
        as.integer()
      y <- dgumbel(seq(0, 100, by = 1), mu = mu_open, sigma = sigma_open)
      y_scaled <- (y - min(y))/(max(y)-min(y))
      p_openness <- y_scaled[x]
      # Now calculate who learns 
      learns <- rbinom(n = length(which(!is.na(population[,1]))), 
                       size = 1, 
                       prob = p_openness)
      # What individuals learned? 
      learners <- which(!is.na(population[,1]))[which(learns == 1)] 
      non_learners <- which(!is.na(population[,1]))[which(learns == 0)]
      # From the learners, who will do so by individual exploration?
      soc_learners <- sample(c(TRUE, FALSE), 
                             length(learners),
                             prob = c(p_soc_learn, (1-p_soc_learn)), 
                             replace = TRUE)
      # Draw random demonstrators 
      demonstrated_traits <- sample(population[which(!is.na(population[,i+1])),i+1], 
                                    size = sum(soc_learners),
                                    replace = T)
      # Now update their beliefs
      # Update beliefs of social learners 
      population[learners[soc_learners],i+2] <- demonstrated_traits
      # Update beliefs of explorers 
      population[learners[!soc_learners],i+2] <- state_world
      # The others just get their previous beliefs
      population[non_learners,i+2] <- population[non_learners,i+1]
      
      #### 2. Surviving ####
      # How many agents are around?
      current_agents <- length(population[which(!is.na(population[,1])),1])
      
      # If there's more than the carrying capacity then start lowering survival probability
      if(current_agents > 100) {
        exponent <- 2/(current_agents/100)
      } else {
        exponent <- 2
      }
      # Calculate the agents probability of survival
      p_survival <- 1 - ((population[which(!is.na(population[,1])),1])/(100-1))^exponent
      # Make sure there are no below-zero probability
      p_survival[p_survival<0] <- 0
      # Which individuals have the correct trait
      fit <- population[which(!is.na(population[,1])),i+2] == state_world
      # Add their bonus 
      p_survival[fit] <- p_survival[fit] + bonus_survival
      # Make sure no probability is higher than 1 
      p_survival[p_survival >=1] <- 1
      # Now calculate who survives 
      survived <- rbinom(n = length(which(!is.na(population[,1]))), 
                         size = 1, 
                         prob = p_survival)
      # NAs in the ages of those who don't
      population[which(!is.na(population[,1]))[survived==0],1] <- NA
      
      
      #### 3. Reproduction ####
      # Pick those in the appropriate age-band
      reproducers <- which(!is.na(population[,1])&
                             population[,1]>=15&
                             population[,1]<=40)
      # From these, who have the appropriate trait?
      fit_reproducers <- population[reproducers,i+2]==state_world
      # Fitness coefficients dependent on carrying capacity
      if (current_agents < 100) {
        fitness_coef <- b
        fitness_coef_fit <- b+bonus_fitness
      } else {
        fitness_coef <- (b/(current_agents/100))
        fitness_coef_fit <- (b+bonus_fitness)/(current_agents/100)
      }
      # Now let's see how many children we get 
      # Fit reproducers
      rf <- rbinom(n = sum(fit_reproducers), 
                   size = 1, 
                   prob = fitness_coef_fit)
      # The rest of reproducers
      r <- rbinom(n = length(reproducers)-sum(fit_reproducers), 
                  size = 1, 
                  prob = fitness_coef)
      # Number of new children 
      num_children <- sum(r) + sum(rf)
      if(num_children > 0) {
        population[(last_agent+1):(last_agent+num_children),1] <- 1
        # They copy reproducers 
        population[(last_agent+1):(last_agent+num_children),i+2] <- 
          sample(population[reproducers,i+2], num_children)
      }
      
      #### 4. Aging ####
      # Everyone ages by one 
      population[which(!is.na(population[,1])),1] <- population[which(!is.na(population[,1])),1]+1
      
      #### 5. World changes #### 
      # The world changes every fifty years
      if (i %in% c(seq(50, t_max, by = 50))) { 
        env_change <-  1 } else {
          env_change  <- 0
        }
      
      # If changes get the current state out and sample from possible worlds
      if (env_change == 1) {
        possible_states <- c(1:4)
        possible_states <- possible_states[-state_world]
        state_world <- sample(possible_states, 1)
      } else {
        state_world <- state_world
      }
      
      #### Book-keeping ####
      df_results[i,1] <- mean(population[,1], na.rm = T)
      df_results[i,2] <- sum(population[which(!is.na(population[,1])),i+2]==1)
      df_results[i,3] <- sum(population[which(!is.na(population[,1])),i+2]==2)
      df_results[i,4] <- sum(population[which(!is.na(population[,1])),i+2]==3)
      df_results[i,5] <- sum(population[which(!is.na(population[,1])),i+2]==4)
      df_results[i,6] <- env_change
      df_results[i,7] <- length(population[which(!is.na(population[,1]))])
      df_results[i,8] <- state_world
    }
    
    #### Results ####
    df <- data.frame(df_results) 
    names(df) <- c("average_age", 
                   "t1",
                   "t2", 
                   "t3", 
                   "t4", 
                   "environment_change",
                   "population", 
                   "state_world")
    df$turn <- 1:nrow(df)
    df$sim <- j 
    
    df <- df %>% 
      mutate(changed = turn*environment_change)
    
    all_results[[j]] <- df
  }
  return(all_results)
}

set.seed(33)

# Low variance
results_low <- varying_windows_model(N = 100, 
                                 t_max = 500, 
                                 b = 0.1, 
                                 bonus_fitness = 0.05, 
                                 bonus_survival = 0.05, 
                                 num_rows = 1e5, 
                                 p_soc_learn = 0.95, 
                                 sigma_open = 2, 
                                 mu_open = 20,
                                 sims = 1)
# Mid variance

results_mid <- varying_windows_model(N = 100, 
                                 t_max = 500, 
                                 b = 0.1, 
                                 bonus_fitness = 0.05, 
                                 bonus_survival = 0.05, 
                                 num_rows = 1e5, 
                                 p_soc_learn = 0.95, 
                                 sigma_open = 5, 
                                 mu_open = 20,
                                 sims = 1)

# high variance

results_high <- varying_windows_model(N = 100, 
                                 t_max = 500, 
                                 b = 0.1, 
                                 bonus_fitness = 0.05, 
                                 bonus_survival = 0.05, 
                                 num_rows = 1e5, 
                                 p_soc_learn = 0.95, 
                                 sigma_open = 10, 
                                 mu_open = 20,
                                 sims = 1)

```

Let's look at what the trajectories of diffusion for each run looked like: 

```{r}
all_runs <- rbind(results_low[[1]], 
                  results_mid[[1]],
                  results_high[[1]])
all_runs <- all_runs %>% 
  mutate(variance = rep(c("low", "mid", "high"), each = 500), 
         variance = ordered(variance, 
                            levels = c("low", 
                                       "mid", 
                                       "high")))

all_runs %>% 
  ggplot(aes(x = turn)) + 
  geom_vline(aes(xintercept = changed), 
             alpha = 0.5, 
             linetype = 2, 
             col = "darkgreen") +
  geom_line(aes(y = t1), col = "darkred") +
  geom_line(aes(y = t2), col = "darkblue") +
  geom_line(aes(y = t3), col = "purple") +
  geom_line(aes(y = t4), col = "orange") +
  theme_bw() + 
  labs(x = "Turn", 
       y = "Prevalence of Traits", 
       title = "Traits through simulation") + 
  facet_wrap(~variance)

```

Right away, we notice differences in the cycles. The high variance run changes a lot more than its counterparts. There is no clear winner in that run, while the low variance simulation tends towards the orange trait, despite environmental changes. 

Let's look at how quickly change happened after the environment changed. 

```{r}
arl <- all_runs %>% 
  mutate(period = rep(c(1:10), 
                      each = 50, 
                      times = 3)) %>% 
  pivot_longer(cols = 2:5, 
               names_to = "trait", 
               values_to = "prevalence") %>% 
  mutate(favored = case_when(trait == "t1" &
                               state_world == 1 ~ 1, 
                             trait == "t2" &
                               state_world == 2 ~ 1,
                             trait == "t3" &
                               state_world == 3 ~ 1,
                             trait == "t4" &
                               state_world == 4 ~ 1))

arl_filtered <- arl %>% 
  filter(favored == 1)

# prepare to label slopes 

out_tidy <- arl_filtered %>% 
  group_by(variance, period) %>% 
  nest() %>% 
  mutate(model = map(data, fit_ols), 
         tidied = map(model, tidy)) %>% 
  unnest(tidied) %>% 
  filter(term == 'turn') %>% 
  select(variance, period, estimate)

arl_filtered <- arl_filtered %>% 
  left_join(out_tidy, by = c('variance', 'period')) 

arl_filtered <- arl_filtered %>% 
  mutate(estimate = as.character(estimate), 
         estimate = as.factor(estimate))

arl_filtered %>% 
  ggplot(aes(x = turn, y = prevalence)) +
  geom_point(col = "orange", 
             size = 1, 
             alpha = 0.5) + 
  geom_smooth(aes(group = period), 
              method = "lm", 
              show.legend = T, 
              se = F, 
              color = 'lightblue', 
              alpha = 0.5) +
  facet_wrap(~variance) + 
  labs(x = "Turn", 
       y = "Prevalence", 
       title = "Prevalence of favored trait at each period")

```

The curves get steeper on average as variance increases. This result is intuitive but I think insightful: as formative periods get longer, we should expect more rapid cultural change. 

Now, I want to see how much consensus there is across the run and how much of that consesus is based on whatever trait happens to be favored in each period. 

```{r}
dominant_df <- arl %>% 
  group_by(variance, turn, trait) %>% 
  summarize(total = sum(prevalence)) %>% 
  arrange(desc(total)) %>% 
  slice(1)

total_pop_df <- arl %>% 
  group_by(variance, turn) %>% 
  summarise(population = population) %>% 
  slice(1)

perc_df <- left_join(dominant_df, 
                     total_pop_df, 
                     by = c("variance", "turn")) %>% 
  mutate(percent_dominant = total/population)
  
perc_df$num_favored <- arl_filtered$prevalence
perc_df$percent_favored <- perc_df$num_favored/ perc_df$total

df_medians <- perc_df %>% 
  group_by(variance) %>% 
  summarise(med = median(percent_dominant))
df_med_favored <- perc_df %>% 
  group_by(variance) %>% 
  summarise(med = mean(percent_favored))

perc_df %>% 
  ggplot(aes(x = turn, y = percent_dominant)) +
  geom_line(col = 'orange') +
  geom_hline(data = df_medians, 
             aes(yintercept = med), 
             col = "red") +
  geom_hline(data = df_med_favored, 
             aes(yintercept = med), 
             col = "black") +
  geom_line(aes(x = turn, y = percent_favored), col = "lightblue", alpha = 0.6) + 
  facet_wrap(~variance) + 
  labs(x = "Turn", 
       y = "% of dominant trait", 
       title = "Consensus across the run")


```

