---
title: "Initial Exploration"
author: "Nicolas Restrepo"
date: "1/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

I am going to use this document to think through the construction of our model. Our main goal is to build a simulation that reflects that agents are more or less open to new information with age. We want to examine how the length and shape of formative periods affects processes of cultural transmission and accumulation. 

This is an early attempt. I am going to be "thinking out loud", and cutting some corners. Hopefully, though, it will give us a good starting point. 

## Set-up 

The first thing is to think about what initial the set-up should look like before anything is put in motion. We need a set of agents, their age, and the cultural traits they acquire and discard. This can be all kept in a data structure we can use to record the agents' characteristics. This is going to be a computationally heavy model so we need to be as economical as possible. Ideally then we want a static structure - no deleting or adding rows or columns. 

I am going to use a matrix with columns equal to the number of turns plus two - one for the age of the agent and one for the initial cultural trait. The number of rows should have to be enough so that it can record all offspring across all rounds. It will then be the number of agents plus the number of agents multiplied by the fertility rate. 

We also need to have traits and environmental change. Let us start this initial exploration by assuming there are four different traits $({t1; t2; t3; t4})$ and four environmental states $({e1; e2; e3; e4})$. Each trait "fits" with their respective environmental state; $t1$ would fit with $e1$. When an individual has the right trait it gives a boost to its baseline survival and reproductive capacity.

The last moving part we need for our set-up is the probability of environmental change. With this, the board is ready and we can start looking at what the agents do each turn. Let's build our setting:

```{r}
#### Set-up ####
# Number of agents
N <- 50
# Maximum turns 
t_max <- 1000
# Baseline fertility rate 
b <- 0.1
# Bonus fertility for fitness
bonus_fitness <- 0.1
# Bonus survival 
bonus_survival <- 0.1
# Repository (enough rows; plus one column for age)
population <- matrix(NA, nrow= N+N*t_max*b+bonus_fitness, 
                     ncol = t_max+1)
# First traits 
population[1:N, 2] <- sample(c(1:4), 
                             size = 50, 
                             replace = T)
# Draw age as integer 
age <- runif(n = N, min = 1, max = 100) %>% as.integer()
# Add to our matrix 
population[1:N,1] <- age
# State world 
state_world <- sample(c(1:4), 1)
# Vector to record if the world changes 
world_changes <- rep(NA, 50)
# Probability of environmental change 
u <- 0.05 

```

To close this section, let me add that all these values are arbitrary and therefore variable. We can - and should - play around with baseline fertility and bonus fertility, as well as with the probability of environmental change. 

## The Model 

Let's think about what a turn would look like. I think a logical sequence would be the following: 

1. Agents learn either individually or socially.
2. Agents either die or survive. 
3. Those who survive and are in reproductive age reproduce.
4. All surviving agents age. 

I am going to look at each step in more detail. 

### Learning 

Right away, we face one of the key questions here:what kind of function is going to determine the shape and length of the agents' formative period. 

My first instinct is that we want newborn agents to be very open so a bell-shaped function - whether Gaussian or not - is not going to be appropriate. I am then going to try an exponential decay function with the following form: 

$$N(t) = N_0e^{-\gamma t}$$

$\gamma$ here controls the rate of decay and could help us explore different shapes. 

```{r}
x <- c(1:100)
y_1 <- exp(1)^((-x)/50)
y_2 <- exp(1)^((-x)/35)
y_3 <- exp(1)^((-x)/25)
y_4 <- exp(1)^((-x)/15)
example_openness <- tibble(age = rep(1:100, 4), 
                      p_o = c(y_1, y_2, y_3, y_4),
                      gamma = as.factor(rep(c(round(1/50,2), round(1/35,2), round(1/25,2),round(1/15,2)), 
                                            each = 100)))
example_openness %>% 
  ggplot(aes(x = age, y = p_o, color = gamma)) +
  geom_line() + 
  theme_bw() + 
  labs(title = "Probability of openness and age")

```

This to me looks plausible - especially values around $\gamma = 0.03$. 25 year-olds have a decent probability of learning at around 0.5, while agents older than 50 have very low probabilities of changing. We will use the probabilities given by this function to see whether agents learn. 

Once we have determined who learns, we need a learning algorithm. I am going to begin with a simple leaning sequence borrowed from McElreath and Deffner. Each turn a learner uses either social learning or individual learning. The probability of social learning is $l$ (here I will say 0.9) and that of individual learning is $1-l$. In social learning, the agent copies a demonstrator at random. For individual learning, they access the state of the environment directly. 

The whole learning sequence would look like this:

```{r}
# These are place markers for the loop 
# Not important right now 
last_agent <- 50
i <- 1
## 1. Learning ##
  # Probability of learning
  x <- population[which(!is.na(population[,1])),1] %>% 
    as.integer()
  p_openness <- exp(1)^((-x)/35)
  # Now calculate who learns 
  learns <- rbinom(n = length(which(!is.na(population[,1]))), 
                   size = 1, 
                   prob = p_openness)
  # What individuals learned? 
  learners <- which(!is.na(population[,1]))[which(learns == 1)] 
  non_learners <- which(!is.na(population[,1]))[which(learns == 0)]
  # Let's see which will learn individually 
  soc_learners <- sample(c(TRUE, FALSE), 
                         length(learners),
                         prob = c(0.9, 0.1), 
                         replace = TRUE)
  # Draw random demonstrators 
  demonstrated_traits <- sample(population[which(!is.na(population[,i+1])),i+1], 
                                size = sum(soc_learners),
                                replace = T)
  # Now update their beliefs
  # Update beliefs of social learners 
  population[learners[soc_learners],i+2] <- demonstrated_traits
  # Update beliefs of explorers 
  population[learners[!soc_learners],i+2] <- state_world
  # The others just get their previous beliefs
  population[non_learners,i+2] <- population[non_learners,i+1]
    
```
Let's run a sanity check to make sure everything worked out. 

```{r}
df_learns <- tibble(age = population[1:50,1], 
                    prob_learning = p_openness,
                    learned = learns, 
                    previous_trait = population[1:50,2], 
                    new_trait = population[1:50, 3])
df_learns %>% head()
```

It seems that it is working as intended. 

### Survival

We need to figure out the relationship between age and the probability of survival. We want a negative relationship but I think it should be non-linear; the rate of decline should be slow at first and get more steep. We could say that it is defined by a power law distribution with negative exponent. Our curve will follow this form: 

$$1-{(\frac{x-a}{b-a})^\alpha}$$

```{r}
x <- c(1:100)
y_1 <- 1 - ((x-1)/(100-1))^1
y_2 <- 1 - ((x-1)/(100-1))^2
y_3 <- 1 - ((x-1)/(100-1))^2.5
y_4 <-1 - ((x-1)/(100-1))^3
example_age <- tibble(age = rep(1:100, 4), 
                      p_s = c(y_1, y_2, y_3, y_4),
                      alpha = as.factor(rep(c(1, 2, 2.5, 3), 
                                            each = 100)))
example_age %>% 
  ggplot(aes(x = age, y = p_s, color = alpha)) +
  geom_line() + 
  theme_bw() + 
  labs(title = "Probability of survival and age")
```

As alpha gets higher, we give younger people more of a chance to survive before old age kicks in. This seems like a plausible starting point for me. 

Let's look at how we would adjudicate the chance of survival each turn. Here, we add a survival bonus for those individuals who have the fitting trait. 

```{r}
## 2. Surviving ##
  # Calculate the agents probability of survival
  p_survival <- 1 - ((population[which(!is.na(population[,1])),1])/(100-1))^2
  # If probability is below zero, turn to 0
  p_survival[p_survival<0] <- 0
  # Which individuals have the correct trait
  fit <- population[which(!is.na(population[,1])),i+2] == state_world
  try(if (sum(is.na(fit)) >0) stop("Wrong sum"))
  # Add their bonus 
  p_survival[fit] <- p_survival[fit] + bonus_survival
  # Make sure no probability is higher than 1 
  p_survival[p_survival >=1] <- 1
  # Now calculate who survives 
  survived <- rbinom(n = length(which(!is.na(population[,1]))), 
                     size = 1, 
                     prob = p_survival)
  # NAs in the ages of those who don't
  population[which(survived == 0),1] <- NA
  
```

Again, let's run a sanity check here to make sure this is doing what it is supposed to do. 

```{r}
df_survives <- tibble(age = population[1:50,1], 
                      p_survival = p_survival, 
                      survived = survived)

head(df_survives)
```

It looks like we it's doing what it's supposed to be doing! Why are some rows NA? Because the agent died so now they have no age. Recall we don't want to dynamically change our data repository because that will make the code slow. NAs will help us keep track of who died without having to remove rows. 

### Reproduction

For reproduction, the decision criteria is more straightforward. We can say that only individuals within a certain age range can reproduce. I am going to assume here that that age group ranges from age 15 to age 40. This range is also something we vary.

Individuals in this age-group will be the group $r$. Agents within that group that also have the environmentally appropriate trait will be the subgroup $r_{fit}$. Agents in $r_{fit}$ will reproduce with probability equal to the baseline reproduction rate plus the fertility bonus. The remaining agents in $r$ will reproduce with probability equal to the baseline fertility rate. 

Each turn, reproduction would look like this: 

```{r}
## 3. Reproducing ##
  # Who are the people in reproductive age 
  reproducers <- which(population[which(!is.na(population[,1]))] >= 15 & 
                         population[which(!is.na(population[,1]))] <= 40)
  # From these, who have the appropriate trait 
  fit_reproducers <- which(population[reproducers,i+2]==state_world)
  # Now let's carry see how many children we get 
  # Fit reproducers
  rf <- rbinom(n = length(fit_reproducers), 
               size = 1, 
               prob = b+bonus_fitness)
  
  # The rest of reproducers
  r <- rbinom(n = length(reproducers)-length(fit_reproducers), 
              size = 1, 
              prob = b)
  # Number of new children 
  num_children <- sum(r) + sum(rf)
  if(num_children > 0) {
    population[(last_agent+1):(last_agent+num_children),1] <- 1
    # They copy reproducers 
    population[(last_agent+1):(last_agent+num_children),i+2] <- 
      sample(population[reproducers,i+2], num_children)
  }
```

The last part of that code just adds the number of children to our matrix in the next available rows. We then sample traits from the demonstrator traits to assign to the newborns. 

### Aging and Environmental change

We only have two pieces of the puzzle left and they are relatively straightforward. We make all surviving agents age by one year. Then, we flip a coin with probability $u$ to see if the environment changes. 

```{r}
## 4. Aging ##
  population[which(!is.na(population[,1])),1] <- population[which(!is.na(population[,1])),1]+1
  
  ## 5. World changes ## 
  # Does it change 
  env_change <- rbinom(n =1, size = 1, prob=u)
  
  # If changes get the current state out and sample from possible worlds
  if (env_change == 1) {
    possible_states <- c(1:4)
    possible_states <- possible_states[-state_world]
    state_world <- sample(possible_states, 1)
    world_changes[i] <- TRUE
  } else {
    state_world <- state_world
    world_changes[i] <- FALSE
  }
```

And that's a simple sketch of our model. Let's see if it is informative at all in this trial run. 

## Making the model go

```{r}
set.seed(33)
#### Set-up ####
# Number of agents
N <- 50
# Maximum turns 
t_max <- 20000
# Baseline fertility rate 
b <- 0.05
# Bonus fertility for fitness
bonus_fitness <- 0.05
# Bonus survival 
bonus_survival <- 0.05
# Repository (enough rows; plus one column for age)
population <- matrix(NA, nrow= N+N*t_max*b+bonus_fitness, 
                     ncol = 250)
# First traits 
population[1:N, 2] <- sample(c(1:4), 
                             size = 50, 
                             replace = T)
# Draw age as integer 
age <- runif(n = N, min = 1, max = 100) %>% as.integer()
# Add to our matrix 
population[1:N,1] <- age
# State world 
state_world <- sample(c(1:4), 1)
# World changes 
world_changes <- rep(NA, 50)
# Probability of environmental change 
u <- 0.05 

# df_results matrix 
df_results <- matrix(NA, nrow = 200, ncol = 7)
#### Loop ####
for(i in 1:200) {
  # Last agent 
  last_agent <- which(!is.na(population[,1]))[length(which(!is.na(population[,1])))]
  ## 1. Learning ##
  # Probability of learning
  x <- population[which(!is.na(population[,1])),1] %>% 
    as.integer()
  p_openness <- exp(1)^((-x)/35)
  # Now calculate who learns 
  learns <- rbinom(n = length(which(!is.na(population[,1]))), 
                   size = 1, 
                   prob = p_openness)
  # What individuals learned? 
  learners <- which(!is.na(population[,1]))[which(learns == 1)] 
  non_learners <- which(!is.na(population[,1]))[which(learns == 0)]
  # Let's see which will learn individually 
  soc_learners <- sample(c(TRUE, FALSE), 
                         length(learners),
                         prob = c(0.9, 0.1), 
                         replace = TRUE)
  # Draw random demonstrators 
  demonstrated_traits <- sample(population[which(!is.na(population[,i+1])),i+1], 
                                size = sum(soc_learners),
                                replace = T)
  # Now update their beliefs
  # Update beliefs of social learners 
  population[learners[soc_learners],i+2] <- demonstrated_traits
  # Update beliefs of explorers 
  population[learners[!soc_learners],i+2] <- state_world
  # The others just get their previous beliefs
  population[non_learners,i+2] <- population[non_learners,i+1]
  
  ## 2. Surviving ##
  # See what the max age is
  # Calculate the agents probability of survival
  p_survival <- 1 - ((population[which(!is.na(population[,1])),1])/(100-1))^2
  p_survival[p_survival<0] <- 0
  # Which individuals have the correct trait
  fit <- population[which(!is.na(population[,1])),i+2] == state_world
  try(if (sum(is.na(fit)) >0) stop("Wrong sum"))
  # Add their bonus 
  p_survival[fit] <- p_survival[fit] + bonus_survival
  # Make sure no probability is higher than 1 
  p_survival[p_survival >=1] <- 1
  # Now calculate who survives 
  survived <- rbinom(n = length(which(!is.na(population[,1]))), 
                     size = 1, 
                     prob = p_survival)
  # NAs in the ages of those who don't
  population[which(survived == 0),1] <- NA
  
  reproducers <- which(!is.na(population[,1])&
                       population[,1]>=15&
                       population[,1]<=40)
# From these, who have the appropriate trait 
fit_reproducers <- which(population[reproducers,i+2]==state_world)
# Now let's carry see how many children we get 
# Fit reproducers
rf <- rbinom(n = length(fit_reproducers), 
             size = 1, 
             prob = b+bonus_fitness)

# The rest of reproducers
r <- rbinom(n = length(reproducers)-length(fit_reproducers), 
            size = 1, 
            prob = b)
# Number of new children 
num_children <- sum(r) + sum(rf)
if(num_children > 0) {
  population[(last_agent+1):(last_agent+num_children),1] <- 1
  # They copy reproducers 
  population[(last_agent+1):(last_agent+num_children),i+2] <- 
    sample(population[reproducers,i+2], num_children)
}

  
  ## 4. Aging ##
  population[which(!is.na(population[,1])),1] <- population[which(!is.na(population[,1])),1]+1
  
  ## 5. World changes ## 
  # Does it change 
  env_change <- rbinom(n =1, size = 1, prob=u)
  
  # If changes get the current state out and sample from possible worlds
  if (env_change == 1) {
    possible_states <- c(1:4)
    possible_states <- possible_states[-state_world]
    state_world <- sample(possible_states, 1)
  } else {
    state_world <- state_world
  }
  table_traits <- table(population[,i+2])
  df_results[i,1] <- mean(population[,1], na.rm = T)
  df_results[i,2] <- table_traits[1]
  df_results[i,3] <- table_traits[2]
  df_results[i,4] <- table_traits[3]
  df_results[i,5] <- table_traits[4]
  df_results[i,6] <- env_change
  df_results[i, 7] <- length(population[which(!is.na(population[,1]))])
}

```

Okay, it runs and it's actually quite efficient. 
let's see the results of our first 200 years. 

```{r}
df <- data.frame(df_results) 
names(df) <- c("average_age", 
                   "t1",
                   "t2", 
                   "t3", 
                   "t4", 
                   "environment_change",
                   "population")
df$turn <- 1:nrow(df)

df <- df %>% 
  mutate(changed = turn*environment_change)


df %>% 
  ggplot(aes(x = turn)) + 
  geom_vline(aes(xintercept = changed), alpha = 0.5, linetype = 2, col = "darkgreen") +
  geom_line(aes(y = t1), col = "darkred") +
  geom_line(aes(y = t2), col = "darkblue") +
  geom_line(aes(y = t3), col = "purple") +
  geom_line(aes(y = t4), col = "orange") +
  theme_bw() + 
  labs(x = "Turn", 
       y = "Prevalence of Traits", 
       title = "Traits through simulation")
```

There are a couple of things that spring to mind here in this first simulation. First, there's a lot of children. This is exemplified by the fact that the population quickly picks up on environmental changes. Maybe a bell-curve to describe openness is not such a bad idea after all. Second, the environment is changing a lot. We want to be able to distinguish cycles driven by environmental change and those driven by cohort replacement. 
